---
title: "Sesión 5: Diseño experimental"
subtitle: "Curso de Análisis de RNA-Seq. Especialización en Bioinformática - Universidad Nacional de Rosario. Argentina"
author:
  - name: "Estefania Mancini"
    orcid: "0000-0002-7308-1273"
    email: "contact@estepi.com"
date: "2026-09"
lang: es
keywords:
  - RNA-Seq
  - Bioconductor
  - Transcriptómica
  - FAIR
  - OpenScience
license: "GNU General Public License v3.0"
repository: "https://github.com/estepi/EB_UNR_2026"
doi: ""
version: "1.0.0"
output:
  html_document:
    theme: flatly
    highlight: tango
    number_sections: true
    code_folding: hide
    code_download: true
    df_print: paged
    fig_caption: yes
    fig_height: 10
    fig_width: 10
    toc: yes
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
  pdf_document:
    toc: yes
    number_sections: true
    latex_engine: xelatex
params:
  fecha_ejecucion: !r Sys.Date()
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Configuración inicial

## Cargamos los paquetes que utilizamos en esta sesión:

  
```{r libraries}
library(DESeq2)
library(org.Mm.eg.db)
```

## Definir el directorio de trabajo


```{r setwd}
setwd("~/Documents/UNR/EB_UNR_2026/")
```

# Exploratory analysis 

## Load metadata

\url{https://www.ncbi.nlm.nih.gov/Traces/study/?acc=PRJNA379959&o=acc_s%3Aa}

```{r loadmd}
md<-read.csv("data/GSE96870_SraRunTable.csv", header = T)
dim(md)
colnames(md)
str(md)
```


## Describe experimental  metadata

```{r exp}
colSums(is.na(md))
```

```{r tissue}
table(md$tissue)
```


```{r infection}
table(md$infection)
```

```{r strain}
table(md$strain)
```

```{r gender}
table(md$gender)
```

```{r Time}
table(md$Time)
```


## Select samples

 The original experimental design in Blackmore et al. 2017 was fairly complex: 
 - 8 week old male and female C57BL/6 mice were collected at Day 0 (before influenza infection), Day 4 and Day 8 after influenza infection. 
 - From each mouse, cerebellum and spinal cord tissues were taken for RNA-seq. 
 - There were originally 4 mice per ‘Sex x Time x Tissue’ group, but a few were lost along the way resulting in a total of 45 samples. For this workshop, we are going to simplify the analysis by only using the 22 cerebellum samples. 
 
 
```{r sel}
selected<-md[md$tissue == "Cerebellum",]
head(selected)
```


# Load counts

- Expression quantification was done using STAR/salmon to align to the mouse genome and then counting reads that map to genes. 


```{r loadcounts}
counts<-read.delim("data/GSE96870_salmon.merged.gene_counts.tsv", header = T, row.names = 1)
counts[1:5,1:5]
counts$gene_name <- NULL
dim(counts)
```

- Columns: samples (total: 22)
- Rows: genes (total: 45706 genes )


```{r round}
counts<-round(counts)
counts[1:5,1:5]
```

# Summarized Experiment

- Create a **SummarizedExperiment** object to store RNA-seq data
- Store count matrix in the assays slot (converted to matrix format)
- Attach sample metadata 

## Match selected samples to counts

```{r selCounts}
ii<- match(colnames(counts), selected$Experiment)
```

## Define metadata 

```{r conditiondf}
coldata <- data.frame(
  sample = colnames(counts),
  Time = selected$Time[ii],
  gender = selected$gender[ii],
  infection = selected$infection[ii],
  mouse_id = selected$mouse_id[ii]
)

rownames(coldata)<-colnames(counts)
coldata$Time<-gsub(" ", "", coldata$Time)
coldata[1:5,]
```


## Check correspondence

```{r checks}
all.equal(rownames(coldata) ,colnames(counts))
```

## Create SummarizedExperiment object

```{r se}
se <- SummarizedExperiment(
    assays  = list(counts = as.matrix(counts)),  
    colData = coldata                            
)
```


## Access counts

```{r assay}
head(assay(se))
```


## Access annotations

```{r ann}
dim(colData(se))
head(colData(se))
```


## Create sample labels 

- combine gender, Time and mouse_id:

```{r label}
se$Label <- paste(se$gender, se$Time, se$mouse_id, sep = "_")
se$Label
colnames(se) <- se$Label
```

## Factorize the Label column to keep in order in plots:

```{r factorlevel, eval=F}
se$Label <- factor(se$Label, levels = se$Label)
```

## Our samples are not in order based on gender and time

```{r order}
se$Group <- paste(se$gender, se$Time, sep = "_")
se$Group
```

## Factorize labels

```{r factorlabel}
se$Group <- factor(se$Group, levels = c("female_Day0","male_Day0", 
                                        "female_Day4","male_Day4",
                                        "female_Day8","male_Day8"))
se <- se[, order(se$Group)]
```


## Save **SummarizedExperiment** object

```{r saveSE}
head(colData(se))
dim(colData(se))

head(assay(se))
dim(assay(se))
    
saveRDS(se, "data/GSE96870_se.rds")

```

## Save options

```{r savothers, eval=FALSE}
savehistory("S5.Rhistory")
```

# Session Info

```{r sessionInfo}
sessionInfo()
```
