---
title: "Sesión 6: Normalización"
subtitle: "Curso de Análisis de RNA-Seq. Especialización en Bioinformática - Universidad Nacional de Rosario. Argentina"
author:
  - name: "Estefania Mancini"
    orcid: "0000-0002-7308-1273"
    email: "contact@estepi.com"
date: "2026-09"
lang: es
keywords:
  - RNA-Seq
  - Bioconductor
  - Transcriptómica
  - FAIR
  - OpenScience
license: "GNU General Public License v3.0"
repository: "https://github.com/estepi/EB_UNR_2026"
doi: ""
version: "1.0.0"
output:
  html_document:
    theme: flatly
    highlight: tango
    number_sections: true
    code_folding: hide
    code_download: true
    df_print: paged
    fig_caption: yes
    fig_height: 10
    fig_width: 10
    toc: yes
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
  pdf_document:
    toc: yes
    number_sections: true
    latex_engine: xelatex
params:
  fecha_ejecucion: !r Sys.Date()
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Configuracion inicial

## Definir el directorio de trabajo


```{r setwd}
setwd("~/Documents/UNR/EB_UNR_2026/")
```
## Cargamos los paquetes que utilizamos en esta sesión:


```{r libraries}
library(DESeq2)
library(SummarizedExperiment)
library(ggplot2)
library(vsn)
library(ComplexHeatmap)
library(RColorBrewer)
library(PCAtools)
```

## Load SE

```{r loadse}
se <- readRDS("data/GSE96870_se.rds")
```


```{r totalgenes}
nrow(se)
```


## Negative binomial distribution

\url{https://hbctraining.github.io/DGE_workshop_salmon/lessons/01_DGE_setup_and_overview.html}

```{r nb}
counts<-assay(se) 

plot0<-ggplot(counts) +
  geom_histogram(aes(x = female_Day0_9), stat = "bin", bins = 200)+
  xlab("Raw expression counts") +
  ylab("Number of genes")+
  theme_bw()

ggsave(filename = "reads0.png" , 
       plot = plot0, 
       width = 16, 
       height = 12, units = "cm")
```
  
  
## Counts distribution

```{r totalgenesRow}
total=sort(rowSums(assay(se, "counts")), decreasing = T)
totalRows<-data.frame(counts=total, 
                      genes=1:length(total))
plot1<-ggplot(totalRows, aes(x=genes,y=counts)) +
geom_point()+
theme_bw()+
   geom_hline(yintercept =5,
              color = "red",
              size=1)
  
###############
ggsave(filename = "reads1.png" , 
       plot = plot1, 
       width = 16, 
       height = 12, units = "cm")
```

```{r totalgenesRow2}
plot2<-ggplot(totalRows, aes(x=genes, y=counts)) +
geom_point()+
theme_bw()+
  ylim(0,50)+
    geom_hline(yintercept =5,
              color = "red",
              size=1)+
    geom_vline(xintercept =25000,
              color = "blue",
              size=1)
###############
ggsave(filename = "reads2.png" , 
       plot = plot2,
       width = 16, 
       height = 12, units = "cm")

```

# by Samples

```{r boxplot}
png("boxplot.png")
boxplot(log10(assay(se)+1),
        ylab=expression('Log'[10]~'Read counts'),
        las=2,
        main="No filter counts")
dev.off()        
```

## Remove genes/rows that do not have > 0 total counts 
- all zero genes

```{r filter0}
se <- se[rowSums(assay(se, "counts")) > 0, ]
nrow(se)
```

## Remove genes/rows that do not have > 5 total counts 

```{r filter5}
se <- se[rowSums(assay(se, "counts")) > 5, ]
nrow(se)
```




## Counts distribution


```{r totalgenesRow3}
total=sort(rowSums(assay(se, "counts")), decreasing = T)

totalRows<-data.frame(counts=round(total), 
                      genes=1:length(total))
plot1<-ggplot(totalRows, aes(x=genes,y=counts)) +
geom_point()+
theme_bw()
###############
ggsave(filename = "reads3.png" , 
       plot = plot1, 
       width = 16, 
       height = 12, units = "cm")
```

```{r totalgenesRow4}
plot2<-ggplot(totalRows, aes(x=genes, y=counts)) +
geom_point()+
theme_bw()+
  ylim(0,50)+
    geom_hline(yintercept =5,
              color = "red",
              size=1)

ggsave(filename = "reads4.png" , 
       plot = plot2,
       width = 16, 
       height = 12, units = "cm")

```

## by Samples

```{r boxplot2}
png("boxplot2.png")

boxplot(log10(assay(se)+1),
        ylab=expression('Log'[10]~'Read counts'),
        las=2,
        main="Abover 5 reads per gene ")
dev.off()        
```

```{r max}
which.max(rowSums(assay(se, "counts")))
```


# Normalization

To perform the median of ratios method of normalization, DESeq2 has a single estimateSizeFactors() function that will generate size factors for us.  This step is automatically performed by the _DESeq()_ function, which we will see later.




```{r deseq2}
dds <- DESeq2::DESeqDataSet(se, design = ~ gender + Time)
dds <- estimateSizeFactors(dds, type="ratio")
se$libSize <-  colSums(assay(se))
```

- get help

```{r estimateSize}
??estimateSizeFactors
```

## Library Sizes

```{r libsize}


plot5<-colData(se) |>
  as.data.frame() |>

  ggplot(aes(x = Label, y = libSize / 1e6, fill = Group)) + 
  geom_bar(stat = "identity") + theme_bw() + 
  labs(x = "Sample", y = "Total count in millions") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

ggsave(filename = "sizeFactor1.png" , 
       plot = plot5,
       width = 16, 
       height = 12, units = "cm")
```

## Plot the size factors against library size and look for any patterns by group:
  
```{r libsizeplot}
plot6<-ggplot(data.frame(libSize = colSums(assay(dds)),
                  sizeFactor = sizeFactors(dds),
                  Group = dds$Group),
       aes(x = libSize, y = sizeFactor, col = Group)) + 
  geom_point(size = 5) + theme_bw() + 
  labs(x = "Library size", y = "Size factor")


ggsave(filename = "sizeFactor2.png" , 
       plot = plot6,
       width = 16, 
       height = 12, units = "cm")
```


# Extract normalized counts by library size:

```{r boxplot3}
cd <- counts(dds,normalized=TRUE)

png("boxplot3.png")
boxplot(log10(as.matrix(cd)+1),ylab=expression('Log'[10]~'Read counts'),las=2,main="DESeq2 norm default")
dev.off()        
```

## Variance stabilization



```{r meanSdPlot}
png("meanSD.png",  width = 480, height = 480, units = "px")
meanSdPlot(assay(dds), ranks = FALSE)
dev.off()  
```

```{r var}

vsd <- DESeq2::vst(dds, blind = TRUE)

png("meanSDpost.png")
meanSdPlot(assay(vsd), ranks = FALSE)
dev.off()  
```

# VST: variance stabilization

```{r ntd}
library("vsn")
ntd <- normTransform(dds)
meanSdPlot(assay(ntd))
```

```{r ddsCV}

vst <- assay(vst(dds))
cv <- as.data.frame(assay(varianceStabilizingTransformation(dds,blind=T)),check.names=F)
```


```{r compareNormVST}
png("compareBoxplots.png",  width = 700, height = 480, units = "px")
par(mfrow=c(1,3))

boxplot(log10(assay(se)+1),
        ylab=expression('Log'[10]~'Read counts'),
        las=2,
        main="Abover 5 reads per gene ")

boxplot(log10(as.matrix(cd)+1),ylab=expression('Log'[10]~'Read counts'),las=2,main="DESeq2 norm default")

boxplot(log10(as.matrix(cv)+1),ylab=expression('Log'[10]~'Read counts'),las=2,main="DESeq2 norm vst")
dev.off()
```


* This function calculates a variance stabilizing transformation (VST) from the fitted dispersion-mean relation(s) and then transforms the count data (normalized by division by the size factors or normalization factors), yielding a matrix of values which are now approximately homoskedastic (having constant variance along the range of mean values).

* The transformation also normalizes with respect to library size. The rlog is less sensitive to size factors, which can be an issue when size factors vary widely. These transformations are useful when checking for outliers or as input for machine learning techniques such as clustering or linear discriminant analysis

* For RNA-seq data, as the mean count value increases, the variance increases. There is a strong almost linear relationship as seen in the figures. The statistical methods such as PCA expects similar variance across the range of mean values. If not, the higher variance genes will contribute more than the lower variance genes. Such data is said to be heteroscedastic and needs to be corrected. Correction using log transformation (with pseudocount) now gives inflates the contribution of the low variance genes. To obtain similar variance across the whole range of mean values.

* As the name suggests, VST transformation stabilizes variance across the whole range of count values. VST is recommended for clustering or visualisation. 

[source](https://hbctraining.github.io/DGE_workshop/lessons/02_DGE_count_normalization.html)

# Check data quality by clustering


## Euclidean distance between samples

```{r dist}
dst <- dist(t(assay(vsd)))

colors <- colorRampPalette(brewer.pal(9, "Blues"))(255)

png("sampleDistances.png")
ComplexHeatmap::Heatmap(
  as.matrix(dst), 
  col = colors,
  name = "Euclidean\ndistance",
  cluster_rows = hclust(dst),
  cluster_columns = hclust(dst),
  bottom_annotation = columnAnnotation(
    gender = vsd$gender,
    time = vsd$Time,
    col = list(gender = c(female = "red", male = "blue"),
               time = c(Day0 = "yellow", 
                        Day4 = "forestgreen", 
                        Day8 = "purple"))))
dev.off()
```

## Principal components analysis

```{r pca}
pcaData <- DESeq2::plotPCA(vsd, intgroup = c("gender", "Time"),
                           returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

png("pcaDef.png")
ggplot(pcaData, aes(x = PC1, y = PC2)) +
  geom_point(aes(color = gender, shape = Time), size = 5) +
  theme_minimal() +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() + 
  scale_color_manual(values = c(male = "blue", female = "red"))
dev.off()
```

## PCA using PCAtools


```{r pcatoolspca}
p <- pca(vst,
         metadata = colData(se), removeVar = 0.1)
```

### Eigenplot

```{r eigenplot}
eigencorplot(p,
    metavars = c('gender',
                 'Time',
                 'infection'))

png("eigenplot.png")
eigencorplot(p,
    metavars = c('gender',
                 'Time',
                 'infection'))
dev.off()
```

### Bi-plots

```{r biplot}
 biplot(p,
    lab = p$metadata$gender,
    hline = 0, vline = 0,
    legendPosition = 'right',
    colby = 'Time')

png("biplotGenderTime.png")
 biplot(p,
    lab = p$metadata$gender,
    hline = 0, vline = 0,
    legendPosition = 'right',
    colby = 'Time')
dev.off()
```

```{r biplot2}

biplot(p,
    lab = p$metadata$Time,
    hline = 0, vline = 0,
    legendPosition = 'right',
    colby = 'gender')

png("biplotTimeGender.png")
biplot(p,
    lab = p$metadata$Time,
    hline = 0, vline = 0,
    legendPosition = 'right',
    colby = 'gender')
dev.off()

```

```{r biplot3}

 biplot(p,
    lab = p$metadata$infection,
    hline = 0, vline = 0,
    legendPosition = 'right',
    colby = 'Time')

png("biplotInfectionTime.png")
 biplot(p,
    lab = p$metadata$infection,
    hline = 0, vline = 0,
    legendPosition = 'right',
    colby = 'Time')
 dev.off()
```

# Save object

```{r saveSE}
head(colData(se))
dim(colData(se))

head(assay(se))
dim(assay(se))
    
saveRDS(se, "data/GSE96870_se.rds")

```

# SessionInfo

```{r sessionInfo}
sessionInfo()
```