---
title: "Sesión 7: Genes diferencialmente expresados"
subtitle: "Curso de Análisis de RNA-Seq. Especialización en Bioinformática - Universidad Nacional de Rosario. Argentina"
author:
  - name: "Estefania Mancini"
    orcid: "0000-0002-7308-1273"
    email: "contact@estepi.com"
date: "2026-09"
lang: es
keywords:
  - RNA-Seq
  - Bioconductor
  - Transcriptómica
  - FAIR
  - OpenScience
license: "GNU General Public License v3.0"
repository: "https://github.com/estepi/EB_UNR_2026"
doi: ""
version: "1.0.0"
output:
  html_document:
    theme: flatly
    highlight: tango
    number_sections: true
    code_folding: hide
    code_download: true
    df_print: paged
    fig_caption: yes
    fig_height: 10
    fig_width: 10
    toc: yes
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
  pdf_document:
    toc: yes
    number_sections: true
    latex_engine: xelatex
params:
  fecha_ejecucion: !r Sys.Date()
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Configuracion inicial

## Definir el directorio de trabajo


```{r setwd}
setwd("~/Documents/UNR/EB_UNR_2026/")
```

## Cargamos los paquetes que utilizamos en esta sesión:

```{r libraries}
library(DESeq2)
library(SummarizedExperiment)
library(ggplot2)
library(vsn)
library(ComplexHeatmap)
library(RColorBrewer)
```

## Load SE

```{r loadse}
se <- readRDS("data/GSE96870_se.rds")
dim(se)
```


# The DESeqDataSet: Create DESeqDataSet

```{r deseq}
dds <- DESeq2::DESeqDataSet(se,
                            design = ~ gender + Time)
```

## Design matrix

```{r modelmatrix}

design(dds)
model.matrix(~ gender + Time , colData(se))


```

```{r modelmatrix0}
model.matrix(~ 0+ gender + Time , colData(se))
```

## Only gender

```{r modelgender}
model.matrix(~ gender , colData(se))
```

# Library sizes and compositional effects

```{r sizefactor}
dds <- estimateSizeFactors(dds)
head(dds$sizeFactor)
head(dds$libSize)
```

# Dispersion

```{r dispersion, message=T}
dds <- estimateDispersions(dds)
```

```{r dispersionplot}
plotDispEsts(dds)

png("plotDisp.png")
plotDispEsts(dds)
dev.off()     

```

# DGE Testing

```{r testing}
dds <- nbinomWaldTest(dds)
```

```{r testingfast, message=TRUE}
ddsFAST <- DESeq(dds)
```


# Explore results 

## Explore results for specific contrasts

### Day 8 vs Day 0

```{r restime}
resTime <- results(dds, 
                   contrast = c("Time", "Day8", "Day0"))
summary(resTime)
class(resTime)

```

```{r reshelp, message=T}
?results
```

## View(resTime)

```{r viewrestime}

head(resTime[order(resTime$pvalue), ])


```


## Plot MA

```{r plotMA}

plotMA(resTime)

png("plotMA.png")
plotMA(resTime)
dev.off()     

```

## LFC = 1

### Day 8 vs Day 0

```{r restimelfc1}
resTimelfc1 <- results(dds, 
                   contrast = c("Time", "Day8", "Day0"),
                    lfcThreshold = 1)
summary(resTimelfc1)
```


```{r plotMAlfc1}


plotMA(resTimelfc1)

png("plotMAlfc1.png")
plotMA(resTimelfc1)
dev.off()     

```


## Shrinkage

```{r shrinkage}
resTimeLfc <- lfcShrink(dds, coef = "Time_Day8_vs_Day0", res = resTime)
```

```{r MAshrinkage}
png("plotMAshrinkage.png")
plotMA(resTimeLfc)
dev.off()
```

## No filtering low expressed

```{r nofilter}
resTimelnofilter <- results(dds, 
                   contrast = c("Time", "Day8", "Day0"),
                    independentFiltering = FALSE          )
summary(resTimelnofilter)
```

```{r plotMAnofilter}
png("plotMAnof.png")
plotMA(resTimelnofilter)
dev.off()
```

# Visualize selected set of genes

## Transform counts

```{r vsd}
vsd <- vst(dds, blind = TRUE)
```


## Get top 10 DE genes

```{r top10}
genes <- resTime[order(resTime$pvalue), ] |>
         head(10) |>
         rownames()
heatmapData <- assay(vsd)[genes, ]
```


## Scale counts for visualization

```{r scale}
heatmapData <- t(scale(t(heatmapData)))
```

## Add annotation

```{r annot}
heatmapColAnnot <- data.frame(colData(vsd)[, c("Time", "gender")])
heatmapColAnnot <- HeatmapAnnotation(df = heatmapColAnnot)
```

## Plot as heatmap

```{r heatmap}
ComplexHeatmap::Heatmap(heatmapData,
                        top_annotation = heatmapColAnnot,
                        cluster_rows = TRUE, cluster_columns = FALSE)

```

```{r plotheatmap}
png("heatmap1.png", width = 700, height = 500, units = "px")
ComplexHeatmap::Heatmap(heatmapData,
                        top_annotation = heatmapColAnnot,
                        cluster_rows = TRUE, cluster_columns = FALSE)
dev.off()
```

# Save results

```{r results}
temp <- as.data.frame(resTime)
write.csv(temp, file = "data/Day8vsDay0.csv")
```

```{r resultsTime}
resTime <- DESeq2::results(dds, contrast = c("Time", "Day8", "Day0"))
timeDE <- as.data.frame(subset(resTime, 
                               padj < 0.05 & abs(log2FoldChange) > log2(1.5)
                       ))
timeDEgenes <- rownames(timeDE)
head(timeDEgenes)
write.csv(timeDEgenes, file="data/timeDEgenes.csv")
```



# Session Info

```{r sessionInfo}
sessionInfo()
```